---
title: "Introduction to Machine Learning"
subtitle: "Final Project"
author: "Group"
date: last-modified
format: 
   html:
     df-print: paged
     embed-resources: true
---

## Description of the Data
The dataset consists of over 500 fine-needle aspiration (FNA) samples.  For each sample, a technician outlined the nuclei from photographs and a software calculated ten characteristics.  Each sample includes the mean, standard error, and maximum of these ten features, resulting in 30 measurements per ID.  The datset includes an unique ID and a label that indicating whether the tissue mass is malignant (M) or benign (B).

## Download the Data

```{r}
#install and load packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
```

```{r}
# download the data, use ID as the row name
FNA_cancer <- read.csv("/Users/ndds/Downloads/FNA_cancer.csv", row.names = 1)
```

## Exploratory Data Analysis
```{r}
# Inspect the data
glimpse(FNA_cancer)
```

```{r}
# review the NA's
colSums(is.na(FNA_cancer))
```

```{r}
#remove the column with all NA's
FNA <- FNA_cancer %>%
  select(-X)
```

```{r}
# change diagnosis to factor
FNA$diagnosis <- as.factor(FNA$diagnosis)
```

```{r}
# create a dummy 1, 0 for diagnosis
FNA <- FNA %>%
  mutate(malignant = ifelse(diagnosis == "M", 1, 0))
```

```{r}
correlation <- FNA %>%
  select(where(is.numeric)) %>%
  cor() %>%
  .["malignant",] %>%
  sort(decreasing = TRUE)

correlation
```

```{r}
# need to create a df for a facet wrap of histograms, use pivot longer
pivot_df <- FNA %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# plot the variables
ggplot(pivot_df, aes(x=values, fill = diagnosis)) +
  geom_histogram(position = "identity") + # overlay the diagnosis
  facet_wrap(~variable, scales = "free_x") # allow x-axis to be independent of each other
```

```{r}
# Review boxplot
ggplot(pivot_df, aes(x = values, fill = diagnosis)) +
  geom_boxplot()+
  facet_wrap(~variable, scales = "free_x")
```
```{r}
# create a mean dataframe
FNA_sub.mean <- FNA %>%
  select(diagnosis, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean, concavity_mean, concave.points_mean, symmetry_mean, fractal_dimension_mean)

# need to create a df for a facet wrap, using pivot longer
pivot_mean <- FNA_sub.mean %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the mean data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#mean
numeric_cols.mean <- sapply(FNA_sub.mean, is.numeric)
corrplot(cor(FNA_sub.mean[,numeric_cols.mean]), method = "circle")
```

```{r}
# create a se dataframe
FNA_sub.se <- FNA %>%
  select(diagnosis, radius_se, texture_se, perimeter_se, area_se, smoothness_se, compactness_se, concavity_se, concave.points_se, symmetry_se, fractal_dimension_se)

# need to create a df for a facet wrap, using pivot longer
pivot_se <- FNA_sub.se %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#se
numeric_cols.se <- sapply(FNA_sub.se, is.numeric)
corrplot(cor(FNA_sub.se[,numeric_cols.se]), method = "circle")
```


```{r}
# create a 'worst' subset dataframe
FNA_sub.worst <- FNA %>%
  select(diagnosis, radius_worst, texture_worst, perimeter_worst, area_worst, smoothness_worst, compactness_worst, concavity_worst, concave.points_worst, symmetry_worst, fractal_dimension_worst, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_worst <- FNA_sub.worst %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_worst, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#worst
numeric_cols.worst <- sapply(FNA_sub.worst, is.numeric)
corrplot(cor(FNA_sub.worst[,numeric_cols.worst]), method = "circle")
```



## Part B

## Part B
