---
title: "Introduction to Machine Learning"
subtitle: "Final Project"
author: "Group"
date: last-modified
format: 
   html:
     df-print: paged
     embed-resources: true
---

## Description of the Data
The dataset consists of over 500 fine-needle aspiration (FNA) samples.  For each sample, a technician outlined the nuclei from photographs and a software calculated ten characteristics.  Each sample includes the mean, standard error, and maximum of these ten features, resulting in 30 measurements per ID.  The datset includes an unique ID and a label that indicating whether the tissue mass is malignant (M) or benign (B).

## Download the Data

```{r}
#install and load packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
```

```{r}
# download the data, use ID as the row name
FNA_cancer <- read.csv("FNA_cancer.csv", row.names = 1)
```

## Exploratory Data Analysis
```{r}
# Inspect the data
glimpse(FNA_cancer)
```

```{r}
# review the NA's
colSums(is.na(FNA_cancer))
```

```{r}
#remove the column with all NA's
FNA <- FNA_cancer %>% dplyr::select(-X)
```

```{r}
# change diagnosis to factor
FNA$diagnosis <- as.factor(FNA$diagnosis)
```

```{r}
# create a dummy 1, 0 for diagnosis
FNA <- FNA %>%
  mutate(malignant = ifelse(diagnosis == "M", 1, 0))
```


# EDA ideas - Anthony - Testing how to use GitHub and add to the project 

```{r}
ggplot(FNA, aes(x=radius_mean, fill=diagnosis)) +
  geom_density(alpha = 0.5)
#showing the clear separation between benign and malignant tumors
```

## Exploratory Data Analysis

#### Inspect and tidy the data

```{r}
# Inspect the data
glimpse(FNA_cancer)
```

#### Correlation Analysis - Correlation of every variable to the variable malignant.

```{r}
# Add a correlation analysis
correlation <- FNA %>%
  dplyr::select(where(is.numeric)) %>%
  cor() %>% # create a correlation matrix
  .["malignant",] %>% # extract the malignant row to find the correlation of all variables to 'malignant' variable
  sort(decreasing = TRUE)

correlation # review
```

-   From the correlation data as it relates to the 'malignant' hot-one encode key, we can see evidence of various strongly correlated predictor variables.

```{r}
# need to create a df for a facet wrap of histograms, use pivot longer
pivot_df <- FNA %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't chosen
               names_to = "variable",
               values_to = "values",)
```

```{r}
# plot the variables
ggplot(pivot_df, aes(x=values, fill = diagnosis)) +
  geom_histogram(position = "identity", bins = 30) + # overlay the diagnosis
  facet_wrap(~variable, scales = "free") # allow x-axis to be independent of each other
```

-   Based on the histograms, some there appears to be better separation with some predictor variables. For example,

```{r}
# Review boxplot
ggplot(pivot_df, aes(x = values, fill = diagnosis)) +
  geom_boxplot() + facet_wrap(~variable, scales = "free_x")
```


```{r}
# create a mean subset dataframe for easier visuals to review
FNA_sub.mean <- FNA %>%
  dplyr::select(diagnosis, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean, concavity_mean, concave.points_mean, symmetry_mean, fractal_dimension_mean, malignant) # select all mean data

# need to create a df for a facet wrap, using pivot longer
pivot_mean <- FNA_sub.mean %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't chosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the mean data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis, fill = diagnosis)) +
  geom_violin()+ # add violinplot
  geom_boxplot(width = 0.2, outlier.shape = NA)+ # overlay a boxplot
  facet_wrap(~variable, scales = "free") +
  theme(legend.position = "none") # no need for legend
```

```{r}
#mean subset correlation plot
numeric_cols.mean <- sapply(FNA_sub.mean, is.numeric)
corrplot(cor(FNA_sub.mean[,numeric_cols.mean]), method = "circle")
```

```{r}
# reviewing the mean data subset histograms
ggplot(pivot_mean, aes(x = values, fill = diagnosis)) +
  geom_histogram(bins = 30)+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
# create a se subset dataframe
FNA_sub.se <- FNA %>%
  dplyr::select(diagnosis, radius_se, texture_se, perimeter_se, area_se, smoothness_se, compactness_se, concavity_se, concave.points_se, symmetry_se, fractal_dimension_se, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_se <- FNA_sub.se %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_se, aes(x = values, y = diagnosis, fill = diagnosis)) +
  geom_violin()+
  geom_boxplot(width = 0.2, outlier.shape = NA)+ # overlay a boxplot
  facet_wrap(~variable, scales = "free") +
  theme(legend.position = "none") # no need for legend
```

```{r}
#se subset correlation plot
numeric_cols.se <- sapply(FNA_sub.se, is.numeric)
corrplot(cor(FNA_sub.se[,numeric_cols.se]), method = "circle")
```

```{r}
# create a 'worst' subset dataframe
FNA_sub.worst <- FNA %>%
  dplyr::select(diagnosis, radius_worst, texture_worst, perimeter_worst, area_worst, smoothness_worst, compactness_worst, concavity_worst, concave.points_worst, symmetry_worst, fractal_dimension_worst, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_worst <- FNA_sub.worst %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't chosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the worst data subset
ggplot(pivot_worst, aes(x = values, y = diagnosis, fill = diagnosis)) +
  geom_violin()+
  geom_boxplot(width = 0.2, outlier.shape = NA)+ # overlay a boxplot
  facet_wrap(~variable, scales = "free")
```

```{r}
#worst subset correlation plot
numeric_cols.worst <- sapply(FNA_sub.worst, is.numeric)
corrplot(cor(FNA_sub.worst[,numeric_cols.worst]), method = "circle")
```

```{r}
#install and load packages
library(rpart)
library(partykit)
library(tidymodels)
library(randomForest)
library(yardstick)
```

## Split the data into test and training data sets

-   Divide the data into a training and a test set. Glimpse the training data.

```{r}
set.seed(1842) # set the seed to 1842

# split the data
FNA_split <- initial_split(FNA,prop=0.8)

# create the training and test data set
FNA_train <- training(FNA_split)
FNA_test <- testing(FNA_split)

# review the training data
glimpse(FNA_train)
```

## Build a classification algorithm using decision trees (prune your tree appropriately)

```{r}
# set a formula
formula_FNA <- as.formula(diagnosis ~ radius_mean + texture_mean + perimeter_mean + area_mean + smoothness_mean + compactness_mean + concavity_mean + concave.points_mean + symmetry_mean + fractal_dimension_mean + radius_se + texture_se + perimeter_se + area_se + smoothness_se + compactness_se + concavity_se + concave.points_se + symmetry_se + fractal_dimension_se + radius_worst + texture_worst +  perimeter_worst + area_worst + smoothness_worst + compactness_worst + concavity_worst +  concave.points_worst + symmetry_worst + fractal_dimension_worst)
```

```{r}
#create a decision tree from the training data set
FNA_Dtree <- rpart(formula_FNA ,data=FNA_train, method = "class")
FNA_Dtree
```

```{r}
# plot the tree
plot(as.party(FNA_Dtree))
```

```{r}
#Calculate the predictions for the test data
FNA_test$preds <- predict(FNA_Dtree,newdata = FNA_test, type = "class")

confusion_dTree <- table(FNA_test$diagnosis,FNA_test$preds, 
                   dnn=c("actual", "predicted"))
confusion_dTree
```

```{r}
# calculate the accuracy
accuracy_tree <- sum(diag(confusion_dTree))/sum(confusion_dTree)
accuracy_tree * 100 # for percentage
```

```{r}
#verifying improvement
printcp(FNA_Dtree) # cross validation error is the best at nplit 4.
```

```{r}
# review cp plot
plotcp(FNA_Dtree)
```

```{r}
# lowest xerror has a cp = 0.01, a prune tree using cp=0.01 will not change anything
new_tree = rpart::prune(FNA_Dtree, cp = 0.01)
# plot the tree
plot(as.party(new_tree))
```

## Build a classification algorithm using random forests or bagging (adjust the parameters of the forest appropriately).

```{r}
p <- length(c("radius_mean", "texture_mean","perimeter_mean","area_mean", "smoothness_mean","compactness_mean","concavity_mean","concave.points_mean", "symmetry_mean","fractal_dimension_mean","radius_se","texture_se","perimeter_se", "area_se","smoothness_se","compactness_se","concavity_se","concave.points_se", "symmetry_se","fractal_dimension_se","radius_worst","texture_worst", "perimeter_worst","area_worst","smoothness_worst","compactness_worst", "concavity_worst","concave.points_worst", "symmetry_worst", "fractal_dimension_worst"))

p
```

-   Model, mtry=5, ntree = 500

```{r}
# set seed for model reproducibility
set.seed(1842)

# m=p, ntree = 500
forest_500 <- randomForest(formula_FNA, data = FNA_train, mtry=5, ntree=500)

#display model
forest_500
```

```{r}
#review the importance
importance(forest_500)

varImpPlot(forest_500, sort = TRUE)
```

-   Using the mtry = 5 model, predict the test data

```{r}
#for reproducibility
set.seed(1842)

# use model to predict
forest_pred<- predict(forest_500,newdata = FNA_test,"class")
```

-   Create a confusion matrix, 0 = Not canceled. 1 = canceled.

```{r}
#make a confusion matrix for the test data 
confusion_mat_forest <- table(forest_pred,FNA_test$diagnosis)
confusion_mat_forest
```

-   Approximately 4.39% missclassified, which aligns with the OOB estimate error.

```{r}
# missclassified
1-sum(diag(confusion_mat_forest))/sum(confusion_mat_forest)
```

## Build a classification algorithm using K-Nearest Neighbors (tune the value of K appropriately)

-   Create the recipe from the formula and training data

```{r}
knn_recipe <- recipe(formula_FNA,data=FNA_train)
```

```{r}
# preprocessing recipe
knn_recipe%>%
  prep()%>% # 
  bake(FNA_train)

# covert hotel to numeric
knn_recipe <- knn_recipe%>%
  step_range(all_numeric_predictors(),min=0,max=1)
```

```{r}
# review the data
knn_prep <- prep(knn_recipe)

bake(knn_prep, new_data = NULL)
```

-   Create the 5 fold cross validation

```{r}
# Create the folds
set.seed(1842)

my_fold <- vfold_cv(FNA_train, v=5)
```

-   Tune the model, set the hyperparameters to tune

```{r}
# using tune(), set neighbors to tune
knn_tune <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn")%>%
  set_mode("classification")
```

-   Add tune to workflow

```{r}
# Add tunable model to workflow
wf_knn_tune <- workflow() %>% 
  add_recipe(knn_recipe) %>%
  add_model(knn_tune)
```

-   Create a grid of k = 3, 5, 7

```{r}
# k equal to 3, 5, and 7
grid_knn <- tibble(neighbors = c(3,5,7))
```

-   Tune a grid, add accuracy, specificity and sensitivity metrics

```{r}
library(yardstick)
# tune
tuned <- tune_grid(wf_knn_tune, 
                   resamples = my_fold, # 5 fold x-validation
                   grid = grid_knn, # add grid for k = 3, 5, 7
                   metrics = metric_set(accuracy, sens, spec)) # add specified metrics

```

-   Display collection metrics for k = 3, 5, 7

```{r}
# review the metrics
collect_metrics(tuned)
```

-   Identify the best k based on accuracy

```{r}
# find the best k based on accuracy
best_k <-select_best(tuned, metric = "accuracy")
best_k
```

-   Use the best k from above to set in the final workflow and fit

```{r}
# set the k at the best identified k
final_wf_knn <- finalize_workflow(wf_knn_tune, best_k)

final_fit <- fit(final_wf_knn, data = FNA_train)
```

-   Provide a confusion matrix

```{r}
augment(new_data=FNA_test, final_fit) %>%
  conf_mat(truth = diagnosis, estimate = .pred_class)
```

-   Provide accuracy, sensitivity, and specificity for this prediction

```{r}
# set the metrics
my_metrics <- metric_set(accuracy, sens, spec)

# list the predictions
predk <- augment(new_data=FNA_test, final_fit) 

# Provide the tibble
my_metrics(predk, truth= diagnosis, estimate = .pred_class)
```

## Build a classification algorithm using Logistic Regression (adjust the model structure appropriately).

### Identify Best Set of Predictors

Due to multicollinearity being present we will use the Step AIC method to determine which subset of predictors from the full model will give the lowest AIC. We have also removed the standard error measurements as they are more suited as information about the mean measurement, not information about the cells themselves. We will use the model recommended by Step AIC for logistic regression.

### Step AIC
```{r}
#| warning: false
formula_FNA_log <- as.formula(diagnosis ~ radius_mean + texture_mean + perimeter_mean + area_mean + smoothness_mean + compactness_mean + concavity_mean + concave.points_mean + symmetry_mean + fractal_dimension_mean + radius_worst + texture_worst +  perimeter_worst + area_worst + smoothness_worst + compactness_worst + concavity_worst +  concave.points_worst + symmetry_worst + fractal_dimension_worst)

mod_full <- glm(formula_FNA_log,data = FNA, family = "binomial")

final_model <- stepAIC(mod_full, trace = 0)
summary(final_model)
```


We can see from the results that we will want a model that only has radius_mean, texture_mean, compactness_mean, concave.points_mean, area_worst, compactness_worst, and symmetry_worst.

### Build Recipe, Model, and Workflow

```{r}

step_AIC_f <- as.formula(diagnosis ~ radius_mean + texture_mean + compactness_mean + 
    concave.points_mean + area_worst + compactness_worst + symmetry_worst)

#Build a tidy models recipe that predicts diagnostics
log_recipe <- recipe(step_AIC_f,data=FNA_train) %>%
  step_normalize(all_numeric_predictors())

#specify the model, logistic regression 
log_reg <- logistic_reg() %>% 
  set_engine("glm") %>%
  set_mode("classification")

# combine in a Workflow
log_wf <- workflow() %>%
  add_model(log_reg)%>%
  add_recipe(log_recipe)
```

### Fit the Model

```{r}
#fit the model
my_fit <- fit(log_wf ,data=FNA_train)

# extract the summary
my_fit%>%
  extract_fit_engine()%>%
  summary()
```


### Calculate Metrics
```{r}
preds_log <- my_fit%>%augment(new_data = FNA_test)
preds_log %>% conf_mat(truth=diagnosis, estimate = .pred_class)
my_metrics <- metric_set(accuracy,sensitivity,specificity)
preds_log %>% my_metrics(truth=diagnosis, estimate = .pred_class)
```


## Create a comparison of the different models on appropriate criteria (accuracy, sensitivity, etc.)






