---
title: "Introduction to Machine Learning"
subtitle: "Final Project"
author: "Group"
date: last-modified
format: 
   html:
     df-print: paged
     embed-resources: true
---

## Description of the Data
The dataset consists of over 500 fine-needle aspiration (FNA) samples.  For each sample, a technician outlined the nuclei from photographs and a software calculated ten characteristics.  Each sample includes the mean, standard error, and maximum of these ten features, resulting in 30 measurements per ID.  The datset includes an unique ID and a label that indicating whether the tissue mass is malignant (M) or benign (B).

## Download the Data

```{r}
#install and load packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
```

```{r}
# download the data, use ID as the row name
FNA_cancer <- read.csv("/Users/ndds/Downloads/FNA_cancer.csv", row.names = 1)
```

## Exploratory Data Analysis
```{r}
# Inspect the data
glimpse(FNA_cancer)
```

```{r}
# review the NA's
colSums(is.na(FNA_cancer))
```

```{r}
#remove the column with all NA's
FNA <- FNA_cancer %>%
  select(-X)
```

```{r}
# change diagnosis to factor
FNA$diagnosis <- as.factor(FNA$diagnosis)
```

```{r}
# create a dummy 1, 0 for diagnosis
FNA <- FNA %>%
  mutate(malignant = ifelse(diagnosis == "M", 1, 0))
```

```{r}
correlation <- FNA %>%
  select(where(is.numeric)) %>%
  cor() %>%
  .["malignant",] %>%
  sort(decreasing = TRUE)

correlation
```

```{r}
# need to create a df for a facet wrap of histograms, use pivot longer
pivot_df <- FNA %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# plot the variables
ggplot(pivot_df, aes(x=values, fill = diagnosis)) +
  geom_histogram(position = "identity") + # overlay the diagnosis
  facet_wrap(~variable, scales = "free_x") # allow x-axis to be independent of each other
```

```{r}
# Review boxplot
ggplot(pivot_df, aes(x = values, fill = diagnosis)) +
  geom_boxplot()+
  facet_wrap(~variable, scales = "free_x")
```
```{r}
# create a mean dataframe
FNA_sub.mean <- FNA %>%
  select(diagnosis, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean, concavity_mean, concave.points_mean, symmetry_mean, fractal_dimension_mean)

# need to create a df for a facet wrap, using pivot longer
pivot_mean <- FNA_sub.mean %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the mean data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#mean
numeric_cols.mean <- sapply(FNA_sub.mean, is.numeric)
corrplot(cor(FNA_sub.mean[,numeric_cols.mean]), method = "circle")
```

```{r}
# create a se dataframe
FNA_sub.se <- FNA %>%
  select(diagnosis, radius_se, texture_se, perimeter_se, area_se, smoothness_se, compactness_se, concavity_se, concave.points_se, symmetry_se, fractal_dimension_se)

# need to create a df for a facet wrap, using pivot longer
pivot_se <- FNA_sub.se %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#se
numeric_cols.se <- sapply(FNA_sub.se, is.numeric)
corrplot(cor(FNA_sub.se[,numeric_cols.se]), method = "circle")
```


```{r}
# create a 'worst' subset dataframe
FNA_sub.worst <- FNA %>%
  select(diagnosis, radius_worst, texture_worst, perimeter_worst, area_worst, smoothness_worst, compactness_worst, concavity_worst, concave.points_worst, symmetry_worst, fractal_dimension_worst, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_worst <- FNA_sub.worst %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_worst, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#worst
numeric_cols.worst <- sapply(FNA_sub.worst, is.numeric)
corrplot(cor(FNA_sub.worst[,numeric_cols.worst]), method = "circle")
```

#EDA ideas - Anthony 

What are we doing? 
We are building a product to improve diagnosis and classify Malignant and Benign tumors.

What features or variables help predict or classify Malignant and Benign tumors?
Can we use patterns? 
Correlation? (corplots are above)
Area, Radius, Concavity - 3 measurements for tumors
Medical procedures - can we use measurements to predict? Do we use KNN? Decision Trees? Random Forests?
We don't want to missclassify or misdiagnos 
Accuracy is key, we DON'T WANT false negatives! That means we miss the diagnosis and the patient suffers. 

Accuracy, Sensitivity, Specificity, Accuracy can help us get false negative ratios (we want these low as possible)

```{r}
ggplot(FNA, aes(x=radius_mean, fill=diagnosis)) +
  geom_density(alpha = 0.5)
#showing the clear separation between benign and malignant tumors
```
I chose Radius, Area, Concavity (3 measurements to identify the shape and size of tumors)

```{r}
#EDA ideas
table(FNA$diagnosis)
#Shows the balance of benign and malignant tumors
```

```{r}
#EDA ideas
prop.table(table(FNA$diagnosis))
```

```{r}
#EDA ideas - these are plots by diagnosis
hist(FNA$radius_mean, main="Radius Mean", col="lightblue")
hist(FNA$area_mean, main="Area Mean", col="lightblue")
hist(FNA$concavity_mean, main="Concavity Mean", col="lightblue")
#these are all right skewed graphs; possibly due to malignant tumors having higher values 
```

```{r}
boxplot(radius_mean ~ diagnosis, data=FNA, col=c("green","red"))
boxplot(area_mean ~ diagnosis, data=FNA, col=c("green","red"))
boxplot(concavity_mean ~ diagnosis, data=FNA, col=c("green","red"))
#normal box plots of the same diagnosis using radius, area, concavity
```

```{r}
boxplot(FNA$area_mean)
```

## Part B

## Part B

### I used the previous clean code to do the Trees and Forests; I don't 100% understand why my plot doesn't show the same output as the original; but we can work through it

#### Train and Split for Trees

```{r}
n_AD <- nrow(FNA)
set.seed(1842)
Test_Index_AD <- sample.int(n_AD,
                            size = round(0.2 * n_AD))
```

```{r}
Train_FNA_AD <- FNA[-Test_Index_AD, ]
Test_FNA <- FNA[Test_Index_AD, ]
```

```{r}
Formula_FNA_Trees_AD <- as.formula(diagnosis ~ .)
```

### Decision Trees

```{r}
FNA_Tree_AD <- rpart(Formula_FNA_Trees_AD,
                     data = Train_FNA_AD,
                     method = "class")
```

```{r}
FNA_Tree_AD
```

```{r}
plot(as.party(FNA_Tree_AD))
```

**I do not know if pruning is necessary because I don't think there is overfitting with diagnosis.**

### Predictions

```{r}
Tree_Predictions_AD <- predict(FNA_Tree_AD,
                               newdata = Test_FNA,
                               type = "class")
```

### Confusion Matrix

```{r}
Confusion_Tree_AD <- table(actual = Test_FNA$diagnosis,
                           predicted = Tree_Predictions_AD)
```

```{r}
Confusion_Tree_AD
```

### Accuracy

```{r}
Tree_Accurary_AD <- sum(diag(Confusion_Tree_AD)) / sum(Confusion_Tree_AD)
```

```{r}
Tree_Accurary_AD
```

### Random Forests

```{r}
set.seed(1842)
FNA_Forest_AD <- randomForest(Formula_FNA_Trees_AD,
                              data = Train_FNA_AD,
                              mtry = 3,
                              ntree = 300,
                              importance = TRUE)
```

```{r}
FNA_Forest_AD
```

```{r}
FNA_Forest_AD$err.rate[300,
                       "OOB"]
```

### Variable Importance

```{r}
importance(FNA_Forest_AD)
```

### Variance Important Plot

```{r}
varImpPlot(FNA_Forest_AD)
```

### Predictions using forests

```{r}
Forest_Prediction_AD <- predict(FNA_Forest_AD,
                                newdata = Test_FNA,
                                type = "class")
```

```{r}
Confusion_Forest_AD <- table(actual = Test_FNA$diagnosis,
                             predicted = Forest_Prediction_AD)
```

```{r}
Confusion_Forest_AD
```

### Accuracy of Forest

```{r}
Forest_Accuracy_AD <- sum(diag(Confusion_Forest_AD)) / sum(Confusion_Forest_AD)
```

```{r}
Forest_Accuracy_AD
```

### Missclass Rate

```{r}
Forest_Missclass_AD <- 1 - Forest_Accuracy_AD
```

```{r}
Forest_Missclass_AD
```
