---
title: "Introduction to Machine Learning"
subtitle: "Final Project"
author: "Group"
date: last-modified
format: 
   html:
     df-print: paged
     embed-resources: true
---

## Description of the Data
The dataset consists of over 500 fine-needle aspiration (FNA) samples.  For each sample, a technician outlined the nuclei from photographs and a software calculated ten characteristics.  Each sample includes the mean, standard error, and maximum of these ten features, resulting in 30 measurements per ID.  The datset includes an unique ID and a label that indicating whether the tissue mass is malignant (M) or benign (B).

## Download the Data

```{r}
#install and load packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
```

```{r}
# download the data, use ID as the row name
FNA_cancer <- read.csv("/Users/ndds/Downloads/FNA_cancer.csv", row.names = 1)
```

## Exploratory Data Analysis
```{r}
# Inspect the data
glimpse(FNA_cancer)
```

```{r}
# review the NA's
colSums(is.na(FNA_cancer))
```

```{r}
#remove the column with all NA's
FNA <- FNA_cancer %>%
  select(-X)
```

```{r}
# change diagnosis to factor
FNA$diagnosis <- as.factor(FNA$diagnosis)
```

```{r}
# create a dummy 1, 0 for diagnosis
FNA <- FNA %>%
  mutate(malignant = ifelse(diagnosis == "M", 1, 0))
```

```{r}
correlation <- FNA %>%
  select(where(is.numeric)) %>%
  cor() %>%
  .["malignant",] %>%
  sort(decreasing = TRUE)

correlation
```

```{r}
# need to create a df for a facet wrap of histograms, use pivot longer
pivot_df <- FNA %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# plot the variables
ggplot(pivot_df, aes(x=values, fill = diagnosis)) +
  geom_histogram(position = "identity") + # overlay the diagnosis
  facet_wrap(~variable, scales = "free_x") # allow x-axis to be independent of each other
```

```{r}
# Review boxplot
ggplot(pivot_df, aes(x = values, fill = diagnosis)) +
  geom_boxplot()+
  facet_wrap(~variable, scales = "free_x")
```
```{r}
# create a mean dataframe
FNA_sub.mean <- FNA %>%
  select(diagnosis, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean, concavity_mean, concave.points_mean, symmetry_mean, fractal_dimension_mean)

# need to create a df for a facet wrap, using pivot longer
pivot_mean <- FNA_sub.mean %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the mean data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#mean
numeric_cols.mean <- sapply(FNA_sub.mean, is.numeric)
corrplot(cor(FNA_sub.mean[,numeric_cols.mean]), method = "circle")
```

```{r}
# create a se dataframe
FNA_sub.se <- FNA %>%
  select(diagnosis, radius_se, texture_se, perimeter_se, area_se, smoothness_se, compactness_se, concavity_se, concave.points_se, symmetry_se, fractal_dimension_se)

# need to create a df for a facet wrap, using pivot longer
pivot_se <- FNA_sub.se %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#se
numeric_cols.se <- sapply(FNA_sub.se, is.numeric)
corrplot(cor(FNA_sub.se[,numeric_cols.se]), method = "circle")
```


```{r}
# create a 'worst' subset dataframe
FNA_sub.worst <- FNA %>%
  select(diagnosis, radius_worst, texture_worst, perimeter_worst, area_worst, smoothness_worst, compactness_worst, concavity_worst, concave.points_worst, symmetry_worst, fractal_dimension_worst, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_worst <- FNA_sub.worst %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_worst, aes(x = values, y = diagnosis)) +
  geom_violin()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
#worst
numeric_cols.worst <- sapply(FNA_sub.worst, is.numeric)
corrplot(cor(FNA_sub.worst[,numeric_cols.worst]), method = "circle")
```

#EDA ideas - Anthony - Testing how to use GitHub and add to the project 

```{r}
ggplot(FNA, aes(x=radius_mean, fill=diagnosis)) +
  geom_density(alpha = 0.5)
#showing the clear separation between benign and malignant tumors
```

## Part B

## Download the Data

```{r}
#install and load packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
```

```{r}
# download the data, use ID as the row name
FNA_cancer <- read.csv("/Users/ndds/Downloads/FNA_cancer.csv", row.names = 1)
```

## Exploratory Data Analysis

#### Inspect and tidy the data

```{r}
# Inspect the data
glimpse(FNA_cancer)
```

-   All data is numeric with a categorical variable for diagnosis and a logical variable for X.

```{r}
# review the NA's
colSums(is.na(FNA_cancer))
```

```{r}
#remove the column with all NA's (X), note the library MASS masks select
FNA <- FNA_cancer %>%
  select(- X)
```

```{r}
# change diagnosis to factor for modeling
FNA$diagnosis <- as.factor(FNA$diagnosis)
```

```{r}
# For EDA, create a dummy 1, 0 for diagnosis
FNA <- FNA %>%
  mutate(malignant = ifelse(diagnosis == "M", 1, 0))
```

#### Correlation Analysis - Correlation of every variable to the variable malignant.

```{r}
# Add a correlation analysis
correlation <- FNA %>%
  select(where(is.numeric)) %>%
  cor() %>% # create a correlation matrix
  .["malignant",] %>% # extract the malignant row to find the correlation of all variables to 'malignant' variable
  sort(decreasing = TRUE)

correlation # review
```

-   From the correlation data as it relates to the 'malignant' hot-one encode key, we can see evidence of various strongly correlated predictor variables.

```{r}
# need to create a df for a facet wrap of histograms, use pivot longer
pivot_df <- FNA %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't chosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# plot the variables
ggplot(pivot_df, aes(x=values, fill = diagnosis)) +
  geom_histogram(position = "identity", bins = 30) + # overlay the diagnosis
  facet_wrap(~variable, scales = "free") # allow x-axis to be independent of each other
```

-   Based on the histograms, some there appears to be better separation with some predictor variables. For example,

```{r}
# Review boxplot
ggplot(pivot_df, aes(x = values, fill = diagnosis)) +
  geom_boxplot()+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
# create a mean subset dataframe for easier visuals to review
FNA_sub.mean <- FNA %>%
  select(diagnosis, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean, concavity_mean, concave.points_mean, symmetry_mean, fractal_dimension_mean, malignant) # select all mean data

# need to create a df for a facet wrap, using pivot longer
pivot_mean <- FNA_sub.mean %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't chosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the mean data subset
ggplot(pivot_mean, aes(x = values, y = diagnosis, fill = diagnosis)) +
  geom_violin()+ # add violinplot
  geom_boxplot(width = 0.2, outlier.shape = NA)+ # overlay a boxplot
  facet_wrap(~variable, scales = "free") +
  theme(legend.position = "none") # no need for legend
```

```{r}
#mean subset correlation plot
numeric_cols.mean <- sapply(FNA_sub.mean, is.numeric)
corrplot(cor(FNA_sub.mean[,numeric_cols.mean]), method = "circle")
```

```{r}
# reviewing the mean data subset histograms
ggplot(pivot_mean, aes(x = values, fill = diagnosis)) +
  geom_histogram(bins = 30)+
  facet_wrap(~variable, scales = "free_x")
```

```{r}
# create a se subset dataframe
FNA_sub.se <- FNA %>%
  select(diagnosis, radius_se, texture_se, perimeter_se, area_se, smoothness_se, compactness_se, concavity_se, concave.points_se, symmetry_se, fractal_dimension_se, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_se <- FNA_sub.se %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't choosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the se data subset
ggplot(pivot_se, aes(x = values, y = diagnosis, fill = diagnosis)) +
  geom_violin()+
  geom_boxplot(width = 0.2, outlier.shape = NA)+ # overlay a boxplot
  facet_wrap(~variable, scales = "free") +
  theme(legend.position = "none") # no need for legend
```

```{r}
#se subset correlation plot
numeric_cols.se <- sapply(FNA_sub.se, is.numeric)
corrplot(cor(FNA_sub.se[,numeric_cols.se]), method = "circle")
```

```{r}
# create a 'worst' subset dataframe
FNA_sub.worst <- FNA %>%
  select(diagnosis, radius_worst, texture_worst, perimeter_worst, area_worst, smoothness_worst, compactness_worst, concavity_worst, concave.points_worst, symmetry_worst, fractal_dimension_worst, malignant)

# need to create a df for a facet wrap, using pivot longer
pivot_worst <- FNA_sub.worst %>%
  pivot_longer(cols = where(is.numeric) & ! diagnosis & ! malignant, # make sure the response isn't chosen
               names_to = "variable",
               values_to = "values")
```

```{r}
# reviewing the worst data subset
ggplot(pivot_worst, aes(x = values, y = diagnosis, fill = diagnosis)) +
  geom_violin()+
  geom_boxplot(width = 0.2, outlier.shape = NA)+ # overlay a boxplot
  facet_wrap(~variable, scales = "free")
```

```{r}
#worst subset correlation plot
numeric_cols.worst <- sapply(FNA_sub.worst, is.numeric)
corrplot(cor(FNA_sub.worst[,numeric_cols.worst]), method = "circle")
```

```{r}
#install and load packages
library(rpart)
library(partykit)
library(tidymodels)
library(randomForest)
library(yardstick)
```

## Split the data into test and training data sets

-   Divide the data into a training and a test set. Glimpse the training data.

```{r}
set.seed(1842) # set the seed to 1842

# split the data
FNA_split <- initial_split(FNA,prop=0.8)

# create the training and test data set
FNA_train <- training(FNA_split)
FNA_test <- testing(FNA_split)

# review the training data
glimpse(FNA_train)
```

## Build a classification algorithm using decision trees (prune your tree appropriately)

```{r}
# set a formula
formula_FNA <- as.formula(diagnosis ~ radius_mean + texture_mean + perimeter_mean + area_mean + smoothness_mean + compactness_mean + concavity_mean + concave.points_mean + symmetry_mean + fractal_dimension_mean + radius_se + texture_se + perimeter_se + area_se + smoothness_se + compactness_se + concavity_se + concave.points_se + symmetry_se + fractal_dimension_se + radius_worst + texture_worst +  perimeter_worst + area_worst + smoothness_worst + compactness_worst + concavity_worst +  concave.points_worst + symmetry_worst + fractal_dimension_worst)
```

```{r}
#create a decision tree from the training data set
FNA_Dtree <- rpart(formula_FNA ,data=FNA_train, method = "class")
FNA_Dtree
```

```{r}
# plot the tree
plot(as.party(FNA_Dtree))
```

```{r}
#Calculate the predictions for the test data
FNA_test$preds <- predict(FNA_Dtree,newdata = FNA_test, type = "class")

confusion_dTree <- table(FNA_test$diagnosis,FNA_test$preds, 
                   dnn=c("actual", "predicted"))
confusion_dTree
```

```{r}
# calculate the accuracy
accuracy_tree <- sum(diag(confusion_dTree))/sum(confusion_dTree)
accuracy_tree * 100 # for percentage
```

```{r}
#verifying improvement
printcp(FNA_Dtree) # cross validation error is the best at nplit 4.
```

```{r}
# review cp plot
plotcp(FNA_Dtree)
```

```{r}
# lowest xerror has a cp = 0.01, a prune tree using cp=0.01 will not change anything
new_tree = rpart::prune(FNA_Dtree, cp = 0.01)
# plot the tree
plot(as.party(new_tree))
```

## Build a classification algorithm using random forests or bagging (adjust the parameters of the forest appropriately).

## Build a classification algorithm using random forests or bagging (adjust the parameters of the forest appropriately).

* Determine the length of predictors
```{r}
p <- length(c("radius_mean", "texture_mean","perimeter_mean","area_mean", "smoothness_mean","compactness_mean","concavity_mean","concave.points_mean", "symmetry_mean","fractal_dimension_mean","radius_se","texture_se","perimeter_se", "area_se","smoothness_se","compactness_se","concavity_se","concave.points_se", "symmetry_se","fractal_dimension_se","radius_worst","texture_worst", "perimeter_worst","area_worst","smoothness_worst","compactness_worst", "concavity_worst","concave.points_worst", "symmetry_worst", "fractal_dimension_worst"))

p
```

* Review options for tree selection in random forest
```{r}
# set seed for model reproducibility
set.seed(1842)

# review OOB error values for different tree sizes
forest_choice <- randomForest(formula_FNA, data = FNA_train, ntree = 1000)

# plot
plot(forest_choice)

# review results
forest_choice$err.rate[300,] # OOB error is slightly better than 500, but not stable around 300
forest_choice$err.rate[500,]
forest_choice$err.rate[800,] # OOB error is good and is very stable
```
* Based on the data above, the OOB error is slightly better than tree at 500, but not stable around tree at 300. The OOB error is better at 800 trees than 500 and is very stable.  Based on this analysis I will review trees at 500 and 800.

* Set the recipe for Tidy Model Random Forest
```{r}
forest_rec <- recipe(formula_FNA,data=FNA_train)
```

* Tidy Model setup, mtry = 5 (sqrt(30) for classifications) and trees = 500
```{r}
#fit a random forest to this data
tidy_forest500 <- rand_forest(mtry = 5, trees=500)%>%
  set_engine("randomForest")%>%
  set_mode("classification")
```

* Build workflow
```{r}
# Add recipe and model to workflow
forest_wf500<- workflow() %>% 
  add_recipe(forest_rec) %>% # add recipe
  add_model(tidy_forest500)
```

* Fit the model
```{r}
#fit the model
set.seed(1842)
forest_fit500 <- fit(forest_wf500,data=FNA_train)
```

* Give the output of this model
```{r}
# the workflow and output
forest_fit500
```

```{r}
# The summary
forest_fit500 %>%
  extract_fit_engine() %>%
  importance()
```

```{r}
# predict the test data for tree
preds_forest500 <- forest_fit500 %>%
  augment(new_data = FNA_test)
```

```{r}
# Confusion matrix and missclassified rate
conf_mat(preds_forest500, truth = diagnosis, .pred_class)
1-accuracy(preds_forest500, truth = diagnosis, .pred_class)$.estimate
```
* Approximately 4.39% miss-classified for the random forest model with 500 tree and mtry at 5.  

```{r}
# Additional metrics
accuracy(preds_forest500, truth = diagnosis, .pred_class)
sens(preds_forest500, truth = diagnosis, .pred_class)
spec(preds_forest500, truth = diagnosis, .pred_class)
roc_auc(preds_forest500, diagnosis, .pred_B)
```

* Try the 800 tree Random Forest Model - Tidy Model setup, mtry = 5 and trees = 800
```{r}
#fit a random forest to this data
tidy_forest800 <- rand_forest(mtry = 5, trees=800)%>%
  set_engine("randomForest")%>%
  set_mode("classification")
```

* Build workflow
```{r}
# Add recipe and model to workflow
forest_wf800<- workflow() %>% 
  add_recipe(forest_rec) %>% # add recipe
  add_model(tidy_forest800)
```

* Fit the model
```{r}
#fit the model
set.seed(1842)
forest_fit800 <- fit(forest_wf800,data=FNA_train)
```

* Give the output of this model
```{r}
# the workflow and output
forest_fit800
```

```{r}
# The summary
forest_fit800 %>%
  extract_fit_engine() %>%
  importance()
```

```{r}
# predict the test data for tree
preds_forest800 <- forest_fit800 %>%
  augment(new_data = FNA_test)
```

```{r}
# Confusion matrix and missclassified rate
conf_mat(preds_forest800, truth = diagnosis, .pred_class)
1-accuracy(preds_forest800, truth = diagnosis, .pred_class)$.estimate
```
* Note - the missclassified rate is the same as the previous model at 4.39%.

```{r}
#metrics for RandomForest at 800 trees
accuracy(preds_forest800, truth = diagnosis, .pred_class)
sens(preds_forest800, truth = diagnosis, .pred_class)
spec(preds_forest800, truth = diagnosis, .pred_class)
roc_auc(preds_forest800, diagnosis, .pred_B)
```

* Review the Bagged model to tune mtry - Tidy Model setup, mtry = p and trees = 500
```{r}
#fit a random forest to this data
tidy_forestBag <- rand_forest(mtry = p, trees=500)%>%
  set_engine("randomForest")%>%
  set_mode("classification")
```

* Build workflow
```{r}
# Add recipe and model to workflow
forest_wfBag<- workflow() %>% 
  add_recipe(forest_rec) %>% # add recipe
  add_model(tidy_forestBag)
```

* Fit the model
```{r}
#fit the model
set.seed(1842)
forest_fitBag <- fit(forest_wfBag,data=FNA_train)
```

* Give the output of this model
```{r}
# the workflow and output
forest_fitBag
```
 *  OOB estimate of  error rate: 4.18% for the bag model compared to OOB estimate of  error rate: 3.74% for ntrees at 500, mtry = 5.  The performance is better with mtry at 5, which could indicate there are highly correlated predictors.  I would use mtry = 5 for my random forest model.


```{r}
# create a ROC for the two models
roc_forest500 <- roc_curve(preds_forest500, diagnosis, .pred_B)

roc_forest800 <- roc_curve(preds_forest800, diagnosis, .pred_B)
```

```{r}
ggplot(data=roc_forest500,aes(x=1-specificity,y=sensitivity,color="Random Forest Model, 500 Trees"))+
  geom_path()+
  geom_abline(color="black")+
  coord_equal() +
  geom_path(data=roc_forest800,aes(x=1-specificity,y=sensitivity,color="Random Forest Model, 800 Trees"))
```

Based on the analysis above, models with 500 and 800 trees (both with mtry as 5) produce the same accuracy and missclassification rate.  Since increasing the number of trees did not improve performance, the model with mtry = 5 and trees = 500 is preferable.  This will help reduce the complexity of the model and decrease the likelihood of overfitting.



## Build a classification algorithm using K-Nearest Neighbors (tune the value of K appropriately)

-   Create the recipe from the formula and training data

```{r}
knn_recipe <- recipe(formula_FNA,data=FNA_train)
```

```{r}
# preprocessing recipe
knn_recipe%>%
  prep()%>% # 
  bake(FNA_train)

# covert hotel to numeric
knn_recipe <- knn_recipe%>%
  step_range(all_numeric_predictors(),min=0,max=1)
```

```{r}
# review the data
knn_prep <- prep(knn_recipe)

bake(knn_prep, new_data = NULL)
```

-   Create the 5 fold cross validation

```{r}
# Create the folds
set.seed(1842)

my_fold <- vfold_cv(FNA_train, v=5)
```

-   Tune the model, set the hyperparameters to tune

```{r}
# using tune(), set neighbors to tune
knn_tune <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn")%>%
  set_mode("classification")
```

-   Add tune to workflow

```{r}
# Add tunable model to workflow
wf_knn_tune <- workflow() %>% 
  add_recipe(knn_recipe) %>%
  add_model(knn_tune)
```

-   Create a grid of k = 3, 5, 7

```{r}
# k equal to 3, 5, and 7
grid_knn <- tibble(neighbors = c(3,5,7))
```

-   Tune a grid, add accuracy, specificity and sensitivity metrics

```{r}
library(yardstick)
# tune
tuned <- tune_grid(wf_knn_tune, 
                   resamples = my_fold, # 5 fold x-validation
                   grid = grid_knn, # add grid for k = 3, 5, 7
                   metrics = metric_set(accuracy, sens, spec)) # add specified metrics

```

-   Display collection metrics for k = 3, 5, 7

```{r}
# review the metrics
collect_metrics(tuned)
```

-   Identify the best k based on accuracy

```{r}
# find the best k based on accuracy
best_k <-select_best(tuned, metric = "accuracy")
best_k
```

-   Use the best k from above to set in the final workflow and fit

```{r}
# set the k at the best identified k
final_wf_knn <- finalize_workflow(wf_knn_tune, best_k)

final_fit <- fit(final_wf_knn, data = FNA_train)
```

-   Provide a confusion matrix

```{r}
augment(new_data=FNA_test, final_fit) %>%
  conf_mat(truth = diagnosis, estimate = .pred_class)
```

-   Provide accuracy, sensitivity, and specificity for this prediction

```{r}
# set the metrics
my_metrics <- metric_set(accuracy, sens, spec)

# list the predictions
predk <- augment(new_data=FNA_test, final_fit) 

# Provide the tibble
my_metrics(predk, truth= diagnosis, estimate = .pred_class)
```

## Build a classification algorithm using logistic regression (adjust the model structure appropriately).

############## need to review more, this has issues

-   Build a tidy models recipe that predicts diagnostics

```{r}
#make a recipe
log_recipe <- recipe(formula_FNA,data=FNA_train)
```

```{r}
# preprocessing recipe
log_recipe%>%
  prep()%>% # 
  bake(FNA_train)

# covert hotel to numeric
log_recipe <- log_recipe%>%
  step_normalize(all_numeric_predictors())
```

-   Build a workflow to fit a logistic regression with your recipe

```{r}
#specify the model, logistic regression 
log_reg <- logistic_reg() %>% 
  set_engine("glm") %>%
  set_mode("classification")

# combine in a Workflow
log_wf <- workflow() %>%
  add_model(log_reg)%>%
  add_recipe(log_recipe)
```

-   Extract the fit

```{r}
#fit the model
my_fit <- fit(log_wf ,data=FNA_train)
my_fit
```

-   Provide a summary of the GLM object.

```{r}
# summary
my_fit%>%
  extract_fit_engine()%>%
  summary()
```

## Create a comparison of the different models on appropriate criteria (accuracy, sensitivity, etc.)
